Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
add Pat data discussion ES pt1

# Hold on to these for now as I update the rest of the rmd.
#demographics <- read.csv("results/tables/denovodemographicsTable.csv")
#OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
#ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
#BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
#ZScoreCombinedData <- read.csv("results/tables/denovoCombinedData.csv")
#ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
#AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
#PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
#ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
#ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
#ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
#ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
#ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
#PowerModelHRR <- read.csv("results/tables/denovoPowerModelHRR.csv")
#PowerModelBFRR <- read.csv("results/tables/denovoPowerModelBFRR.csv")
#SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
#SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
#RRSimNNeededH <- read.csv("results/tables/denovoRRSimNNeededH.csv")
#RRSimNNeededBF <- read.csv("results/tables/denovoRRSimNNeededBF.csv")

@@

```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

#data1 order Escobar, Baxter, Ross, HMP, Arumugam, Wu, Goodrich, Zupancic

ZH <- as.numeric(ZScoreCombinedData$ZH)
bmi <- as.numeric(ZScoreCombinedData$BMI)
BMI.class <- as.character(ZScoreCombinedData$BMICat)
BMI.class[BMI.class == "Normal"] <- "Lean"
obese <- as.character(ZScoreCombinedData$Obese)
ZlogBF <- as.numeric(ZScoreCombinedData$ZLogBF)

data1 <- as.data.frame(cbind(ZH, ZlogBF, bmi))

#Create group analysis table to harness full data set
#For Z-score normalization
t.obese <- as.character(ZScoreTurnbaughData$Obese)
t.bmi.class <- as.character(ZScoreTurnbaughData$BMICat)
t.ZH <- ZScoreTurnbaughData$ZH
t.ZlogBF <- ZScoreTurnbaughData$ZLogBF

#Combine everything together
comb1.obese <- c(obese, t.obese)
comb1.bmi.class <- c(BMI.class, t.bmi.class)
comb1.ZH <- as.data.frame(c(ZH, t.ZH))
group.data.test1 <- cbind(comb1.ZH, comb1.obese, comb1.bmi.class)
colnames(group.data.test1) <- c("ZH", "obese", "BMI.class")

#Let's change the order so it makes sense
group.data.test1$BMI.class <- factor(group.data.test1$BMI.class, levels = c("Lean", "Overweight", "Obese"))

group.data.test <- group.data.test1

#Bacteroidetes/Firmicutes Datasets
comb.ZlogBF <- as.data.frame(c(ZlogBF, t.ZlogBF))
group.data.test <- cbind(comb.ZlogBF, group.data.test)
colnames(group.data.test)[1] <- "ZLogBF"

## Perform the statistical tests
combZHTtest <- t.test(ZH ~ obese, data=group.data.test) #P-value = 0.006492
combZHANOVA <- anova(lm(ZH ~ BMI.class, data=group.data.test)) #P-value = 0.01588
a2 <- aov(ZH ~ BMI.class, data=group.data.test)
combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
combZHTukeyFinal <- combZHTukey$BMI.class

combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.6337
combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.7125
a2 <- aov(ZLogBF ~ BMI.class, data=group.data.test)
combZLogBFTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
combZLogBFTukeyFinal <- combZLogBFTukey$BMI.class

@@

Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obesityy could be observed. Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.  Similarily, we found that the other two approaches, that used the pooled data approach, was also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4]. For these approaches both Shannon diversity and B/F ratio failed to have a significant correlation to obesity. In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.  Overall we found that the most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported differences between studies ranged from `r format(round(min(ActualDiff$H), 2), nsmall=2)`-`r format(round(max(ActualDiff$H), 2), nsmall=2)` percent.  Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.  Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B/F ratio [Figure 5B & 6B].  Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B/F ratio.

>>>>

#Using a quick for loop to generate the necessary effect size information
variableTested <- seq(1:6)
variables <- colnames(alpha_power.obs)[1:6]
alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
studies <- rownames(beta_tests.summary)

for(i in variableTested){
  tempdata <- c()
  for(j in studies){
  tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
  }
  alpha_effectSize[, i] <- tempdata
}

colnames(alpha_effectSize) <- variables
rownames(alpha_effectSize) <- studies

@@

@@

Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obesityy could be observed. Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.  Similarily, we found that the other two approaches, that used the pooled data approach, was also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4]. For these approaches both Shannon diversity and B/F ratio failed to have a significant correlation to obesity. In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.  Overall we found that most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported effect sizes between studies ranged from `r format(round(min(alpha_effectSize[, 'shannon']), 2), nsmall=2)`-`r format(round(max(alpha_effectSize[, 'shannon']), 2), nsmall=2)` percent.  Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.  Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B/F ratio [Figure 5B & 6B].  Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B/F ratio.
